---
title: "Hotel Booking Demand Analysis"
author: "Alireza Samea"
date: "2026-04-04"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
## Project Overview

This analysis explores the [Hotel Booking Demand dataset](https://www.kaggle.com/datasets/jessemostipak/hotel-booking-demand), 
a publicly available dataset on Kaggle containing 119,390 hotel booking records across two hotel types: 
city hotels and resort hotels. The data includes information on booking patterns, guest demographics, 
pricing, and cancellations.

The central business question we are trying to answer is: **What factors influence hotel booking 
cancellations, pricing, and guest behavior — and what can hotel managers do about it?**

**Dataset:** Jesse Mostipak via Kaggle | 119,390 rows | 32 columns

---

## Analysis Plan

1. **Data Loading and Exploration** — import libraries, load data, inspect structure
2. **Data Cleaning and Preparation** — handle missing values, fix data types, create new variables
3. **Exploratory Data Analysis (EDA)** — booking patterns, seasonal trends, hotel type comparisons
4. **Hypothesis Testing** — inferential statistics to test key business assumptions
5. **Cancellation Analysis** — what predicts whether a booking gets cancelled
6. **Key Insights and Recommendations** — summary of findings for hotel management

---

## Section 1: Data Loading and Exploration
### 1.1 Load Libraries

We will use the following packages:

- **tidyverse**: a collection of R packages for data science...
- **knitr**: formats tables and output neatly in the HTML document
```{r}
library(tidyverse)
library(knitr)
```

### 1.2 Load the Dataset

We load the CSV file into a data frame called `hotel_data`. A data frame is R's version of a table 
where rows are observations (individual bookings) and columns are variables (features of each booking).
```{r}
hotel_data <- read.csv("Dataset/hotel_bookings.csv")
dim(hotel_data)
head(hotel_data)
```

The dataset contains **119,390 rows** and **32 columns**, meaning we have 119,390 individual hotel 
bookings with 32 pieces of information recorded for each one.

## Section 2: Data Cleaning and Preparation

Before we analyze the data, we need to inspect and clean it. Raw datasets often contain missing values, 
incorrect data types, and variables that need to be transformed before they are useful for analysis.

### 2.1 Inspect the Data

We start by checking the column names and structure of the dataset. The `colnames()` function shows us 
all 32 variable names, and `str()` shows us the data type of each column along with a preview of its values.
```{r}
colnames(hotel_data)
str(hotel_data)
```

The output shows us that some columns are stored as integers when they should be treated as categories. 
For example, `is_canceled` is coded as 0 and 1 but represents a yes/no outcome, not a number. 
We will fix these in the next step.

### 2.2 Check for Missing Values

We check how many missing values exist in each column using `colSums(is.na())`.
This helps us decide how to handle incomplete data before analysis.

However, this dataset stores missing values as the string `"NULL"` instead of leaving cells blank.
We can confirm this by running `class(hotel_data$agent)` which returns `"character"`, and 
`head(hotel_data$agent)` which shows the literal text `"NULL"` in those cells. This means R does 
not automatically recognize them as missing values. We fix this by reloading the data with the 
`na.strings` argument, which tells R to treat `"NULL"` as `NA`.
```{r}
hotel_data <- read.csv("Dataset/hotel_bookings.csv", na.strings = c("", "NA", "NULL"))
colSums(is.na(hotel_data))
```
Four columns contain missing values. The `agent` and `company` columns have the most, with 16,340 
and 112,593 missing values respectively. This makes business sense because not every booking is made 
through a travel agent or a corporate account. We will handle these in the next step.

### 2.3 Handle Missing Values

We handle each column differently based on what the missing value actually means in context.
For `children`, `agent`, and `company`, a missing value means zero — no children, no agent, 
and no corporate account. For `country`, we leave the 488 missing values as is since we cannot 
make a safe assumption about where a guest is from.
```{r}
hotel_data$children[is.na(hotel_data$children)] <- 0
hotel_data$agent[is.na(hotel_data$agent)]       <- 0
hotel_data$company[is.na(hotel_data$company)]   <- 0

colSums(is.na(hotel_data))
```
After handling missing values, only the 488 missing entries in `country` remain, which we will 
exclude only if a specific analysis requires country-level filtering.

### 2.4 Fix Data Types

Some columns are stored as integers when they actually represent categories. We convert them to 
factors so R treats them correctly in analysis and visualizations. We also fix `arrival_date_month` 
which is stored as text but sorts alphabetically by default. We force it into the correct calendar 
order using an ordered factor.
```{r}
hotel_data$is_canceled <- factor(hotel_data$is_canceled, 
                                  levels = c(0, 1), 
                                  labels = c("No", "Yes"))

hotel_data$is_repeated_guest <- factor(hotel_data$is_repeated_guest, 
                                        levels = c(0, 1), 
                                        labels = c("No", "Yes"))

hotel_data$arrival_date_month <- factor(hotel_data$arrival_date_month,
                                         levels = c("January", "February", "March",
                                                    "April", "May", "June", "July",
                                                    "August", "September", "October",
                                                    "November", "December"),
                                         ordered = TRUE)

str(hotel_data[c("is_canceled", "is_repeated_guest", "arrival_date_month")])
```

The `str()` output confirms the three columns are now stored as factors with the correct levels 
and ordering.

### 2.5 Summary Statistics

Before checking for specific issues, we run a statistical summary of the entire dataset. 
This gives us a quick overview of min, max, mean, and quartiles for every column, and is 
often the fastest way to spot suspicious values.
```{r}
summary(hotel_data)
```

The summary reveals a few issues we will address: `adr` has a minimum value of -6.38 which 
is impossible for a room rate, and some bookings show zero guests across all guest type columns 
which is logically impossible. We will clean these in section 2.7.

### 2.6 Check for Duplicates

We check for duplicate rows to ensure each booking record is unique. Duplicate rows can 
skew counts and summaries if left in the data.
```{r}
sum(duplicated(hotel_data))
```

The output shows 31,994 duplicate rows. We will remove them to ensure our analysis is based 
on unique booking records.
```{r}
hotel_data <- hotel_data[!duplicated(hotel_data), ]
dim(hotel_data)
```

After removing duplicates, the dataset contains 87,396 rows.

### 2.7 Standardize Text Values

We convert the `hotel` column to lowercase and trim any extra spaces to ensure 
consistent values throughout the analysis.
```{r}
hotel_data$hotel <- trimws(tolower(hotel_data$hotel))
unique(hotel_data$hotel)
```
The output confirms we have exactly two clean hotel types: `"city hotel"` and `"resort hotel"`.


