---
title: "Hotel Booking Demand Analysis"
author: "Alireza Samea"
date: "2026-04-04"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Project Overview

This analysis explores the [Hotel Booking Demand dataset](https://www.kaggle.com/datasets/jessemostipak/hotel-booking-demand), 
a publicly available dataset on Kaggle containing 119,390 hotel booking records across two hotel types: 
city hotels and resort hotels. The data includes information on booking patterns, guest demographics, 
pricing, and cancellations.

The central business question we are trying to answer is: **What factors influence hotel booking 
cancellations, pricing, and guest behavior and what can hotel managers do about it?**

**Dataset:** Jesse Mostipak via Kaggle | 119,390 rows | 32 columns

## Analysis Plan

1. **Data Loading and Exploration** - import libraries, load data, inspect structure
2. **Data Cleaning and Preparation** - handle missing values, fix data types, create new variables
3. **Exploratory Data Analysis (EDA)** - booking patterns, seasonal trends, hotel type comparisons
4. **Hypothesis Testing** - inferential statistics to test key business assumptions
5. **Cancellation Analysis** - what predicts whether a booking gets cancelled
6. **Key Insights and Recommendations** - summary of findings for hotel management

## Section 1: Data Loading and Exploration

### 1.1 Load Libraries

We will use the following packages:

- **tidyverse**: a collection of R packages for data science, including `ggplot2` for visualization, 
  `dplyr` for data manipulation, and `readr` for importing data
- **knitr**: formats tables and output neatly in the HTML document
```{r}
library(tidyverse)
library(knitr)
```

### 1.2 Load the Dataset

We load the CSV file into a data frame called `hotel_data`. A data frame is R's version of a table 
where rows are observations (individual bookings) and columns are variables (features of each booking).
```{r}
hotel_data <- read.csv("Dataset/hotel_bookings.csv")
dim(hotel_data)
head(hotel_data)
```

The dataset contains **119,390 rows** and **32 columns**, meaning we have 119,390 individual hotel 
bookings with 32 pieces of information recorded for each one. Note that we will reload the data 
in Section 2.2 with proper settings after inspecting its structure.

## Section 2: Data Cleaning and Preparation

Before we analyze the data, we need to inspect and clean it. Raw datasets often contain missing values, 
incorrect data types, and variables that need to be transformed before they are useful for analysis.

### 2.1 Inspect the Data

We start by checking the column names and structure of the dataset. The `colnames()` function shows us 
all 32 variable names, and `str()` shows us the data type of each column along with a preview of its values.
```{r}
colnames(hotel_data)
str(hotel_data)
```

The output shows us that some columns are stored as integers when they should be treated as categories. 
For example, `is_canceled` is coded as 0 and 1 but represents a yes/no outcome, not a number. 
We will fix these in the next step.

### 2.2 Check for Missing Values

We check how many missing values exist in each column using `colSums(is.na())`.
This helps us decide how to handle incomplete data before analysis.

However, this dataset stores missing values as the string `"NULL"` instead of leaving cells blank.
We can confirm this by running `class(hotel_data$agent)` which returns `"character"`, and 
`head(hotel_data$agent)` which shows the literal text `"NULL"` in those cells. This means R does 
not automatically recognize them as missing values. We fix this by reloading the data with the 
`na.strings` argument, which tells R to treat `"NULL"` as `NA`.
```{r}
hotel_data <- read.csv("Dataset/hotel_bookings.csv", na.strings = c("", "NA", "NULL"))
colSums(is.na(hotel_data))
```

Four columns contain missing values. The `agent` and `company` columns have the most, with 16,340 
and 112,593 missing values respectively. This makes business sense because not every booking is made 
through a travel agent or a corporate account. We will handle these in the next step.

### 2.3 Handle Missing Values

We handle each column differently based on what the missing value actually means in context.
For `children`, `agent`, and `company`, a missing value means zero - no children, no agent, 
and no corporate account. For `country`, we leave the 488 missing values as is since we cannot 
make a safe assumption about where a guest is from.
```{r}
hotel_data$children[is.na(hotel_data$children)] <- 0
hotel_data$agent[is.na(hotel_data$agent)]       <- 0
hotel_data$company[is.na(hotel_data$company)]   <- 0

colSums(is.na(hotel_data))
```

After handling missing values, only the 488 missing entries in `country` remain, which we will 
exclude only if a specific analysis requires country-level filtering.

### 2.4 Fix Data Types

Some columns are stored as integers when they actually represent categories. We convert them to 
factors so R treats them correctly in analysis and visualizations. We also fix `arrival_date_month` 
which is stored as text but sorts alphabetically by default. We force it into the correct calendar 
order using an ordered factor.
```{r}
hotel_data$is_canceled <- factor(hotel_data$is_canceled, 
                                  levels = c(0, 1), 
                                  labels = c("No", "Yes"))

hotel_data$is_repeated_guest <- factor(hotel_data$is_repeated_guest, 
                                        levels = c(0, 1), 
                                        labels = c("No", "Yes"))

hotel_data$arrival_date_month <- factor(hotel_data$arrival_date_month,
                                         levels = c("January", "February", "March",
                                                    "April", "May", "June", "July",
                                                    "August", "September", "October",
                                                    "November", "December"),
                                         ordered = TRUE)

str(hotel_data[c("is_canceled", "is_repeated_guest", "arrival_date_month")])
```

The `str()` output confirms the three columns are now stored as factors with the correct levels 
and ordering.

### 2.5 Summary Statistics

Before checking for specific issues, we run a statistical summary of the entire dataset. 
This gives us a quick overview of min, max, mean, and quartiles for every column, and is 
often the fastest way to spot suspicious values.
```{r}
summary(hotel_data)
```

The summary reveals a few issues we will address: `adr` has a minimum value of -6.38 which 
is impossible for a room rate, and some bookings show zero guests across all guest type columns 
which is logically impossible. We will clean these in section 2.8.

### 2.6 Check for Duplicates

We check for duplicate rows to ensure each booking record is unique. Duplicate rows can 
skew counts and summaries if left in the data.
```{r}
sum(duplicated(hotel_data))
```

The output shows 31,994 rows that are identical across all 32 columns. Since the dataset 
does not contain a unique booking ID column, we cannot confirm with certainty that these 
are data entry errors rather than legitimate identical bookings. However, having two bookings 
that are identical across all 32 variables simultaneously is statistically unlikely, so we 
assume these are duplicates and remove them.
```{r}
hotel_data <- hotel_data[!duplicated(hotel_data), ]
dim(hotel_data)
```

After removing assumed duplicates, the dataset contains 87,396 rows. This assumption should 
be noted as a limitation of this analysis.

### 2.7 Standardize Text Values

We convert the `hotel` column to lowercase and trim any extra spaces to ensure 
consistent values throughout the analysis.
```{r}
hotel_data$hotel <- trimws(tolower(hotel_data$hotel))
unique(hotel_data$hotel)
```

The output confirms we have exactly two clean hotel types: `"city hotel"` and `"resort hotel"`.

### 2.8 Remove Suspicious Values

The summary revealed that `adr` has a minimum value of -6.38 which is impossible for a room 
rate. The summary also shows bookings where `adults`, `children`, and `babies` are all zero, 
which is logically impossible. We remove both types of suspicious records.
```{r}
hotel_data <- hotel_data[hotel_data$adr >= 0, ]

hotel_data <- hotel_data[!(hotel_data$adults == 0 & 
                            hotel_data$children == 0 & 
                            hotel_data$babies == 0), ]
dim(hotel_data)
```

After removing suspicious values the dataset is clean and ready for analysis.

### 2.9 Create New Variables

We create several new variables that will make our analysis and visualizations cleaner.
```{r}
hotel_data$stay_length <- hotel_data$stays_in_weekend_nights + 
                          hotel_data$stays_in_week_nights

hotel_data$total_guests <- hotel_data$adults + 
                           hotel_data$children + 
                           hotel_data$babies

hotel_data$arrival_date <- as.Date(paste(hotel_data$arrival_date_year,
                                         hotel_data$arrival_date_month,
                                         hotel_data$arrival_date_day_of_month,
                                         sep = "-"), format = "%Y-%B-%d")

hotel_data$family <- ifelse(hotel_data$children + hotel_data$babies > 0, 
                            "Family", "Non-Family")

summary(hotel_data[c("stay_length", "total_guests", "arrival_date", "family")])
```

These new variables will be used throughout the EDA and hypothesis testing sections 
to make comparisons cleaner and more meaningful.

## Section 3: Exploratory Data Analysis

In this section we explore the data visually to understand booking patterns, seasonal trends, 
and differences between hotel types before moving into formal hypothesis testing.

### 3.1 Booking Volume by Hotel Type

We start by comparing the total number of bookings between city hotels and resort hotels 
to understand the overall split in the dataset.
```{r}
ggplot(hotel_data, aes(x = hotel, fill = hotel)) +
  geom_bar() +
  labs(title = "Booking Volume by Hotel Type",
       x = "Hotel Type",
       y = "Number of Bookings") +
  theme_minimal() +
  theme(legend.position = "none")
```

City hotels account for roughly twice the number of bookings compared to resort hotels, 
suggesting they are the more dominant hotel type in this dataset.

### 3.2 Bookings by Month

We look at how bookings are distributed across the months of the year to identify 
seasonal patterns. Note that months are displayed in correct calendar order because 
we converted `arrival_date_month` to an ordered factor in Section 2.4.
```{r}
ggplot(hotel_data, aes(x = arrival_date_month, fill = arrival_date_month)) +
  geom_bar() +
  labs(title = "Booking Volume by Month",
       x = "Month",
       y = "Number of Bookings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

Bookings peak during the summer months of July and August and reach their lowest point 
in January and February, reflecting a clear seasonal demand pattern typical in the 
hospitality industry.

### 3.3 Cancellation Rate by Hotel Type

We compare the proportion of cancelled bookings between city hotels and resort hotels. 
Understanding which hotel type has a higher cancellation rate helps managers plan 
overbooking strategies and resource allocation.
```{r}
ggplot(hotel_data, aes(x = hotel, fill = is_canceled)) +
  geom_bar(position = "fill") +
  labs(title = "Cancellation Rate by Hotel Type",
       x = "Hotel Type",
       y = "Proportion",
       fill = "Cancelled") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
```

City hotels have a noticeably higher cancellation rate compared to resort hotels, suggesting that city hotel bookings are more likely to be speculative or subject to last minute changes.

### 3.4 Average Daily Rate by Hotel Type

We compare the average daily rate (ADR) between city hotels and resort hotels using a boxplot. ADR represents the average revenue earned per occupied room per day and is one of the most important pricing metrics in the hospitality industry.
```{r}
ggplot(hotel_data, aes(x = hotel, y = adr, fill = hotel)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 500)) +
  labs(title = "Average Daily Rate by Hotel Type",
       x = "Hotel Type",
       y = "Average Daily Rate (ADR)") +
  theme_minimal() +
  theme(legend.position = "none")
```
Note: The y axis is capped at 500 for readability. A small number of extreme outliers above this threshold exist in the data but are excluded from the display to avoid compressing the main distribution.

Resort hotels show a wider spread in ADR compared to city hotels, suggesting more variable pricing likely driven by seasonality. City hotels have a more consistent and slightly higher median ADR.

### 3.5 Average Daily Rate by Month

We look at how ADR changes across months to understand seasonal pricing patterns. 
This helps hotel managers understand when demand and revenue are highest.
```{r}
ggplot(hotel_data, aes(x = arrival_date_month, y = adr, fill = arrival_date_month)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 500)) +
  labs(title = "Average Daily Rate by Month",
       x = "Month",
       y = "Average Daily Rate (ADR)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

ADR peaks during the summer months of July and August, consistent with the higher booking volumes we saw in Section 3.2. This confirms that summer is the high season where hotels can charge premium prices.

### 3.6 Cancellation Rate by Month

We examine whether cancellation rates vary by month. A high cancellation rate in certain months could significantly impact revenue even when booking volumes are high.
```{r}
ggplot(hotel_data, aes(x = arrival_date_month, fill = is_canceled)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cancellation Rate by Month",
       x = "Month",
       y = "Proportion",
       fill = "Cancelled") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Cancellation rates are fairly consistent across all months, hovering between 20% and 35%. There is a slight increase in cancellations during spring and early summer months (March through August), while January, November, and December show the lowest cancellation rates.

### 3.7 Lead Time Distribution by Hotel Type

Lead time is the number of days between the booking date and the arrival date. We compare lead time distributions between hotel types to understand how far in advance guests typically book each hotel type.
```{r}
ggplot(hotel_data, aes(x = lead_time, fill = hotel)) +
  geom_histogram(binwidth = 10, alpha = 0.7, position = "identity") +
  coord_cartesian(xlim = c(0, 365)) +
  labs(title = "Lead Time Distribution by Hotel Type",
       x = "Lead Time (days)",
       y = "Number of Bookings",
       fill = "Hotel Type") +
  theme_minimal()
```
Note: The x axis is capped at 365 days for readability. A small number of bookings with lead times beyond one year exist in the data but are excluded from the display.

Both hotel types show a right-skewed distribution with most bookings made close to the arrival date. City hotels show a slightly longer tail, suggesting a higher proportion of far-in-advance bookings compared to resort hotels.

### 3.8 Cancellation Rate by Deposit Type

We examine whether the type of deposit made at booking affects cancellation rates. Intuitively, guests who pay a non-refundable deposit should be less likely to cancel.
```{r}
ggplot(hotel_data, aes(x = deposit_type, fill = is_canceled)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cancellation Rate by Deposit Type",
       x = "Deposit Type",
       y = "Proportion",
       fill = "Cancelled") +
  theme_minimal()
```

Surprisingly, non-refundable deposits show the highest cancellation rate. This is counterintuitive but may reflect that non-refundable bookings are often made far in advance through online travel agents, giving guests more time to change plans even at the cost of losing their deposit.

### 3.9 Lead Time by Cancellation Status

We examine whether cancelled bookings tend to have longer lead times. Guests who book far in advance may be more likely to cancel as their plans change over time.
```{r}
ggplot(hotel_data, aes(x = is_canceled, y = lead_time, fill = is_canceled)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 365)) +
  labs(title = "Lead Time by Cancellation Status",
       x = "Cancelled",
       y = "Lead Time (days)",
       fill = "Cancelled") +
  theme_minimal() +
  theme(legend.position = "none")
```

Cancelled bookings have a noticeably higher median lead time compared to bookings that were not cancelled. This confirms that guests who book further in advance are more likely to cancel, which is an important insight for hotel revenue management strategies.

















