---
title: "Hotel Booking Demand Analysis"
author: "Alireza Samea"
date: "2026-04-04"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Project Overview

This analysis explores the [Hotel Booking Demand dataset](https://www.kaggle.com/datasets/jessemostipak/hotel-booking-demand), 
a publicly available dataset on Kaggle containing 119,390 hotel booking records across two hotel types: 
city hotels and resort hotels. The data includes information on booking patterns, guest demographics, 
pricing, and cancellations.

The central business question we are trying to answer is: **What factors influence hotel booking 
cancellations, pricing, and guest behavior and what can hotel managers do about it?**

**Dataset:** Jesse Mostipak via Kaggle | 119,390 rows | 32 columns

## Analysis Plan

1. **Data Loading and Exploration** - import libraries, load data, inspect structure
2. **Data Cleaning and Preparation** - handle missing values, fix data types, create new variables
3. **Exploratory Data Analysis (EDA)** - booking patterns, seasonal trends, hotel type comparisons
4. **Hypothesis Testing** - inferential statistics to test key business assumptions
5. **Cancellation Analysis** - what predicts whether a booking gets cancelled
6. **Key Insights and Recommendations** - summary of findings for hotel management

## Section 1: Data Loading and Exploration

### 1.1 Load Libraries

We will use the following packages:

- **tidyverse**: a collection of R packages for data science, including `ggplot2` for visualization, 
  `dplyr` for data manipulation, and `readr` for importing data
- **knitr**: formats tables and output neatly in the HTML document
```{r}
library(tidyverse)
library(knitr)
```

### 1.2 Load the Dataset

We load the CSV file into a data frame called `hotel_data`. A data frame is R's version of a table 
where rows are observations (individual bookings) and columns are variables (features of each booking).
```{r}
hotel_data <- read.csv("Dataset/hotel_bookings.csv")
dim(hotel_data)
head(hotel_data)
```

The dataset contains **119,390 rows** and **32 columns**, meaning we have 119,390 individual hotel 
bookings with 32 pieces of information recorded for each one. Note that we will reload the data 
in Section 2.2 with proper settings after inspecting its structure.

## Section 2: Data Cleaning and Preparation

Before we analyze the data, we need to inspect and clean it. Raw datasets often contain missing values, 
incorrect data types, and variables that need to be transformed before they are useful for analysis.

### 2.1 Inspect the Data

We start by checking the column names and structure of the dataset. The `colnames()` function shows us 
all 32 variable names, and `str()` shows us the data type of each column along with a preview of its values.
```{r}
colnames(hotel_data)
str(hotel_data)
```

The output shows us that some columns are stored as integers when they should be treated as categories. 
For example, `is_canceled` is coded as 0 and 1 but represents a yes/no outcome, not a number. 
We will fix these in the next step.

### 2.2 Check for Missing Values

We check how many missing values exist in each column using `colSums(is.na())`.
This helps us decide how to handle incomplete data before analysis.

However, this dataset stores missing values as the string `"NULL"` instead of leaving cells blank.
We can confirm this by running `class(hotel_data$agent)` which returns `"character"`, and 
`head(hotel_data$agent)` which shows the literal text `"NULL"` in those cells. This means R does 
not automatically recognize them as missing values. We fix this by reloading the data with the 
`na.strings` argument, which tells R to treat `"NULL"` as `NA`.
```{r}
hotel_data <- read.csv("Dataset/hotel_bookings.csv", na.strings = c("", "NA", "NULL"))
colSums(is.na(hotel_data))
```

Four columns contain missing values. The `agent` and `company` columns have the most, with 16,340 
and 112,593 missing values respectively. This makes business sense because not every booking is made 
through a travel agent or a corporate account. We will handle these in the next step.

### 2.3 Handle Missing Values

We handle each column differently based on what the missing value actually means in context.
For `children`, `agent`, and `company`, a missing value means zero - no children, no agent, 
and no corporate account. For `country`, we leave the 488 missing values as is since we cannot 
make a safe assumption about where a guest is from.
```{r}
hotel_data$children[is.na(hotel_data$children)] <- 0
hotel_data$agent[is.na(hotel_data$agent)]       <- 0
hotel_data$company[is.na(hotel_data$company)]   <- 0

colSums(is.na(hotel_data))
```

After handling missing values, only the 488 missing entries in `country` remain, which we will 
exclude only if a specific analysis requires country-level filtering.

### 2.4 Fix Data Types

Some columns are stored as integers when they actually represent categories. We convert them to 
factors so R treats them correctly in analysis and visualizations. We also fix `arrival_date_month` 
which is stored as text but sorts alphabetically by default. We force it into the correct calendar 
order using an ordered factor.
```{r}
hotel_data$is_canceled <- factor(hotel_data$is_canceled, 
                                  levels = c(0, 1), 
                                  labels = c("No", "Yes"))

hotel_data$is_repeated_guest <- factor(hotel_data$is_repeated_guest, 
                                        levels = c(0, 1), 
                                        labels = c("No", "Yes"))

hotel_data$arrival_date_month <- factor(hotel_data$arrival_date_month,
                                         levels = c("January", "February", "March",
                                                    "April", "May", "June", "July",
                                                    "August", "September", "October",
                                                    "November", "December"),
                                         ordered = TRUE)

str(hotel_data[c("is_canceled", "is_repeated_guest", "arrival_date_month")])
```

The `str()` output confirms the three columns are now stored as factors with the correct levels 
and ordering.

### 2.5 Summary Statistics

Before checking for specific issues, we run a statistical summary of the entire dataset. 
This gives us a quick overview of min, max, mean, and quartiles for every column, and is 
often the fastest way to spot suspicious values.
```{r}
summary(hotel_data)
```

The summary reveals a few issues we will address: `adr` has a minimum value of -6.38 which 
is impossible for a room rate, and some bookings show zero guests across all guest type columns 
which is logically impossible. We will clean these in section 2.8.

### 2.6 Check for Duplicates

We check for duplicate rows to ensure each booking record is unique. Duplicate rows can 
skew counts and summaries if left in the data.
```{r}
sum(duplicated(hotel_data))
```

The output shows 31,994 rows that are identical across all 32 columns. Since the dataset 
does not contain a unique booking ID column, we cannot confirm with certainty that these 
are data entry errors rather than legitimate identical bookings. However, having two bookings 
that are identical across all 32 variables simultaneously is statistically unlikely, so we 
assume these are duplicates and remove them.
```{r}
hotel_data <- hotel_data[!duplicated(hotel_data), ]
dim(hotel_data)
```

After removing assumed duplicates, the dataset contains 87,396 rows. This assumption should 
be noted as a limitation of this analysis.

### 2.7 Standardize Text Values

We convert the `hotel` column to lowercase and trim any extra spaces to ensure 
consistent values throughout the analysis.
```{r}
hotel_data$hotel <- trimws(tolower(hotel_data$hotel))
unique(hotel_data$hotel)
```

The output confirms we have exactly two clean hotel types: `"city hotel"` and `"resort hotel"`.

### 2.8 Remove Suspicious Values

The summary revealed that `adr` has a minimum value of -6.38 which is impossible for a room 
rate. The summary also shows bookings where `adults`, `children`, and `babies` are all zero, 
which is logically impossible. We remove both types of suspicious records.
```{r}
hotel_data <- hotel_data[hotel_data$adr >= 0, ]

hotel_data <- hotel_data[!(hotel_data$adults == 0 & 
                            hotel_data$children == 0 & 
                            hotel_data$babies == 0), ]
dim(hotel_data)
```

After removing suspicious values the dataset is clean and ready for analysis.

### 2.9 Create New Variables

We create several new variables that will make our analysis and visualizations cleaner.
```{r}
hotel_data$stay_length <- hotel_data$stays_in_weekend_nights + 
                          hotel_data$stays_in_week_nights

hotel_data$total_guests <- hotel_data$adults + 
                           hotel_data$children + 
                           hotel_data$babies

hotel_data$arrival_date <- as.Date(paste(hotel_data$arrival_date_year,
                                         hotel_data$arrival_date_month,
                                         hotel_data$arrival_date_day_of_month,
                                         sep = "-"), format = "%Y-%B-%d")

hotel_data$family <- ifelse(hotel_data$children + hotel_data$babies > 0, 
                            "Family", "Non-Family")

summary(hotel_data[c("stay_length", "total_guests", "arrival_date", "family")])
```

These new variables will be used throughout the EDA and hypothesis testing sections 
to make comparisons cleaner and more meaningful.

## Section 3: Exploratory Data Analysis

In this section we explore the data visually to understand booking patterns, seasonal trends, 
and differences between hotel types before moving into formal hypothesis testing.

### 3.1 Booking Volume by Hotel Type

We start by comparing the total number of bookings between city hotels and resort hotels 
to understand the overall split in the dataset.
```{r}
ggplot(hotel_data, aes(x = hotel, fill = hotel)) +
  geom_bar() +
  labs(title = "Booking Volume by Hotel Type",
       x = "Hotel Type",
       y = "Number of Bookings") +
  theme_minimal() +
  theme(legend.position = "none")
```

City hotels account for roughly twice the number of bookings compared to resort hotels, 
suggesting they are the more dominant hotel type in this dataset.

### 3.2 Bookings by Month

We look at how bookings are distributed across the months of the year to identify 
seasonal patterns. Note that months are displayed in correct calendar order because 
we converted `arrival_date_month` to an ordered factor in Section 2.4.
```{r}
ggplot(hotel_data, aes(x = arrival_date_month, fill = arrival_date_month)) +
  geom_bar() +
  labs(title = "Booking Volume by Month",
       x = "Month",
       y = "Number of Bookings") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

Bookings peak during the summer months of July and August and reach their lowest point 
in January and February, reflecting a clear seasonal demand pattern typical in the 
hospitality industry.

### 3.3 Cancellation Rate by Hotel Type

We compare the proportion of cancelled bookings between city hotels and resort hotels. 
Understanding which hotel type has a higher cancellation rate helps managers plan 
overbooking strategies and resource allocation.
```{r}
ggplot(hotel_data, aes(x = hotel, fill = is_canceled)) +
  geom_bar(position = "fill") +
  labs(title = "Cancellation Rate by Hotel Type",
       x = "Hotel Type",
       y = "Proportion",
       fill = "Cancelled") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()
```

City hotels have a noticeably higher cancellation rate compared to resort hotels, suggesting that city hotel bookings are more likely to be speculative or subject to last minute changes.

### 3.4 Average Daily Rate by Hotel Type

We compare the average daily rate (ADR) between city hotels and resort hotels using a boxplot. ADR represents the average revenue earned per occupied room per day and is one of the most important pricing metrics in the hospitality industry.
```{r}
ggplot(hotel_data, aes(x = hotel, y = adr, fill = hotel)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 500)) +
  labs(title = "Average Daily Rate by Hotel Type",
       x = "Hotel Type",
       y = "Average Daily Rate (ADR)") +
  theme_minimal() +
  theme(legend.position = "none")
```
Note: The y axis is capped at 500 for readability. A small number of extreme outliers above this threshold exist in the data but are excluded from the display to avoid compressing the main distribution.

Resort hotels show a wider spread in ADR compared to city hotels, suggesting more variable pricing likely driven by seasonality. City hotels have a more consistent and slightly higher median ADR.

### 3.5 Average Daily Rate by Month

We look at how ADR changes across months to understand seasonal pricing patterns. 
This helps hotel managers understand when demand and revenue are highest.
```{r}
ggplot(hotel_data, aes(x = arrival_date_month, y = adr, fill = arrival_date_month)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 500)) +
  labs(title = "Average Daily Rate by Month",
       x = "Month",
       y = "Average Daily Rate (ADR)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

ADR peaks during the summer months of July and August, consistent with the higher booking volumes we saw in Section 3.2. This confirms that summer is the high season where hotels can charge premium prices.

### 3.6 Cancellation Rate by Month

We examine whether cancellation rates vary by month. A high cancellation rate in certain months could significantly impact revenue even when booking volumes are high.
```{r}
ggplot(hotel_data, aes(x = arrival_date_month, fill = is_canceled)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cancellation Rate by Month",
       x = "Month",
       y = "Proportion",
       fill = "Cancelled") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Cancellation rates are fairly consistent across all months, hovering between 20% and 35%. There is a slight increase in cancellations during spring and early summer months (March through August), while January, November, and December show the lowest cancellation rates.

### 3.7 Lead Time Distribution by Hotel Type

Lead time is the number of days between the booking date and the arrival date. We compare lead time distributions between hotel types to understand how far in advance guests typically book each hotel type.
```{r}
ggplot(hotel_data, aes(x = lead_time, fill = hotel)) +
  geom_histogram(binwidth = 10, alpha = 0.7, position = "identity") +
  coord_cartesian(xlim = c(0, 365)) +
  labs(title = "Lead Time Distribution by Hotel Type",
       x = "Lead Time (days)",
       y = "Number of Bookings",
       fill = "Hotel Type") +
  theme_minimal()
```
Note: The x axis is capped at 365 days for readability. A small number of bookings with lead times beyond one year exist in the data but are excluded from the display.

Both hotel types show a right-skewed distribution with most bookings made close to the arrival date. City hotels show a slightly longer tail, suggesting a higher proportion of far-in-advance bookings compared to resort hotels.

### 3.8 Cancellation Rate by Deposit Type

We examine whether the type of deposit made at booking affects cancellation rates. Intuitively, guests who pay a non-refundable deposit should be less likely to cancel.
```{r}
ggplot(hotel_data, aes(x = deposit_type, fill = is_canceled)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Cancellation Rate by Deposit Type",
       x = "Deposit Type",
       y = "Proportion",
       fill = "Cancelled") +
  theme_minimal()
```

Surprisingly, non-refundable deposits show the highest cancellation rate. This is counterintuitive but may reflect that non-refundable bookings are often made far in advance through online travel agents, giving guests more time to change plans even at the cost of losing their deposit.

### 3.9 Lead Time by Cancellation Status

We examine whether cancelled bookings tend to have longer lead times. Guests who book far in advance may be more likely to cancel as their plans change over time.
```{r}
ggplot(hotel_data, aes(x = is_canceled, y = lead_time, fill = is_canceled)) +
  geom_boxplot() +
  coord_cartesian(ylim = c(0, 365)) +
  labs(title = "Lead Time by Cancellation Status",
       x = "Cancelled",
       y = "Lead Time (days)",
       fill = "Cancelled") +
  theme_minimal() +
  theme(legend.position = "none")
```

Cancelled bookings have a noticeably higher median lead time compared to bookings that were not cancelled. This confirms that guests who book further in advance are more likely to cancel, which is an important insight for hotel revenue management strategies.

## Section 4: Hypothesis Testing

In this section we use inferential statistics to test whether the patterns we observed in the EDA are statistically significant or could simply be due to random chance. We will conduct two hypothesis tests based on questions raised by the EDA.

### 4.1 Is There a Significant Difference in Mean Lead Time Between Hotel Types?

From Section 3.7 we observed a difference in lead time distributions between hotel types. 
We now test whether this difference is statistically significant.

**Null Hypothesis (H0):** The mean lead time for city hotels is equal to the mean lead 
time for resort hotels.

**Alternative Hypothesis (H1):** The mean lead time for city hotels is not equal to the 
mean lead time for resort hotels.

We use a two-tailed Welch two-sample t-test because we are not assuming a direction, 
and we do not assume equal variances between the two groups. We verify the variances 
are different before proceeding:
```{r}
city_hotels   <- hotel_data[hotel_data$hotel == "city hotel", ]
resort_hotels <- hotel_data[hotel_data$hotel == "resort hotel", ]

var(city_hotels$lead_time)
var(resort_hotels$lead_time)
```

The variances are substantially different, confirming that Welch is the appropriate choice over a standard two-sample t-test.
```{r}
t.test(city_hotels$lead_time, resort_hotels$lead_time, 
       var.equal = FALSE, 
       alternative = "two.sided")
```

The p-value is far below 0.05, so we reject the null hypothesis. There is strong statistical evidence that lead times differ between hotel types. Interestingly, resort hotels show a slightly higher mean lead time (83.4 days) compared to city hotels (77.8 days), suggesting guests plan resort stays further in advance than city stays.

### 4.2 Is the Average Length of Stay Significantly Different From 3 Nights?

From Section 2.9 we saw that the average stay length is approximately 3.6 nights. We now test whether this is significantly different from 3 nights or whether the difference could be due to random chance.

**Null Hypothesis (H0):** The mean length of stay is equal to 3 nights.

**Alternative Hypothesis (H1):** The mean length of stay is not equal to 3 nights.

We use a one-sample t-test comparing the sample mean against the hypothesized value of 3. This is a two-tailed test because we are asking whether the mean is different in either direction, not specifically higher or lower.
```{r}
t.test(hotel_data$stay_length, mu = 3, alternative = "two.sided")
```

The p-value is far below 0.05, so we reject the null hypothesis. The average length of stay is significantly different from 3 nights. The sample mean of approximately 3.6 nights suggests guests tend to stay slightly longer than 3 nights on average. Hotels should factor this into staffing schedules, housekeeping planning, and extended stay pricing strategies.

### 4.3 Do Cancelled Bookings Have a Higher Mean Lead Time?

From Section 3.9 we observed that cancelled bookings tend to have longer lead times. We now test whether this difference is statistically significant.

**Null Hypothesis (H0):** The mean lead time for cancelled bookings is equal to the mean lead time for non-cancelled bookings.

**Alternative Hypothesis (H1):** The mean lead time for cancelled bookings is greater than the mean lead time for non-cancelled bookings.

We use a one-tailed Welch t-test because we have a clear directional hypothesis based on our EDA observation in Section 3.9.
```{r}
cancelled     <- hotel_data[hotel_data$is_canceled == "Yes", ]
not_cancelled <- hotel_data[hotel_data$is_canceled == "No", ]

t.test(cancelled$lead_time, not_cancelled$lead_time,
       var.equal = FALSE,
       alternative = "greater")
```

The p-value is far below 0.05, so we reject the null hypothesis. Cancelled bookings have a significantly higher mean lead time than non-cancelled bookings. This confirms that guests who book further in advance are more likely to cancel, which is a critical insight for hotel revenue management and overbooking strategies.

### 4.4 Is the Mean ADR Significantly Different Between Hotel Types?

From Section 3.4 we observed that city hotels appear to have a slightly higher and more consistent ADR compared to resort hotels. We now test whether this difference is statistically significant.

**Null Hypothesis (H0):** The mean ADR for city hotels is equal to the mean ADR for resort hotels.

**Alternative Hypothesis (H1):** The mean ADR for city hotels is not equal to the mean ADR for resort hotels.

We use a two-tailed Welch t-test because we are not assuming a direction, and the boxplot in Section 3.4 showed clearly different spreads between the two groups, confirming unequal variances.
```{r}
t.test(city_hotels$adr, resort_hotels$adr,
       var.equal = FALSE,
       alternative = "two.sided")
```

The p-value is far below 0.05, so we reject the null hypothesis. There is strong statistical evidence that mean ADR differs significantly between hotel types. City hotels have a higher mean ADR, suggesting they command premium pricing likely driven by their urban location and business traveler demand.

### 4.5 Do Family Bookings Have a Longer Mean Stay Than Non-Family Bookings?

From our `family` variable created in Section 2.9 we can test whether families tend to stay longer than non-family guests. This is an intuitive business question since families typically travel for leisure and may stay longer than business travelers.

**Null Hypothesis (H0):** The mean length of stay for family bookings is equal to the mean length of stay for non-family bookings.

**Alternative Hypothesis (H1):** The mean length of stay for family bookings is greater than the mean length of stay for non-family bookings.

We use a one-tailed Welch t-test because we have a clear directional hypothesis — families are expected to stay longer, not just differently.
```{r}
family_bookings     <- hotel_data[hotel_data$family == "Family", ]
non_family_bookings <- hotel_data[hotel_data$family == "Non-Family", ]

t.test(family_bookings$stay_length, non_family_bookings$stay_length,
       var.equal = FALSE,
       alternative = "greater")
```

The p-value is far below 0.05, so we reject the null hypothesis. Family bookings have a significantly longer mean stay than non-family bookings. Hotels should consider targeted packages for families such as extended stay discounts or family room bundles to capitalize on this pattern.

## Section 5: Cancellation Analysis

In this section we go beyond visualization to quantify cancellation patterns and identify which factors are the strongest predictors of whether a booking will be cancelled. We use logistic regression because our outcome variable `is_canceled` is binary — either Yes or No.

### 5.1 Overall Cancellation Rate

We start by establishing the baseline cancellation rate across the entire dataset.
```{r}
cancellation_rate <- prop.table(table(hotel_data$is_canceled))
round(cancellation_rate * 100, 2)
```

Approximately 27.5% of all bookings in the cleaned dataset were cancelled. This is notably high and confirms that cancellation is a significant business problem worth investigating further.

### 5.2 Cancellation Rate by Key Variables

We calculate exact cancellation rates by deposit type, hotel type, and customer type to identify which segments have the highest cancellation risk.
```{r}
# Cancellation rate by deposit type
round(prop.table(table(hotel_data$deposit_type, hotel_data$is_canceled), margin = 1) * 100, 2)

# Cancellation rate by hotel type
round(prop.table(table(hotel_data$hotel, hotel_data$is_canceled), margin = 1) * 100, 2)

# Cancellation rate by customer type
round(prop.table(table(hotel_data$customer_type, hotel_data$is_canceled), margin = 1) * 100, 2)
```

The tables confirm the pattern we saw in Section 3.8 — non-refundable deposit bookings have a surprisingly high cancellation rate. City hotels cancel at a higher rate than resort hotels, and transient customers show the highest cancellation rate among customer types.

### 5.3 Logistic Regression Model

We build a logistic regression model to identify which variables are the strongest 
predictors of cancellation. Logistic regression is appropriate here because our outcome 
variable is binary — cancelled or not cancelled.
```{r}
model <- glm(is_canceled ~ lead_time + deposit_type + hotel + 
               total_guests + stays_in_week_nights + customer_type,
             data = hotel_data,
             family = binomial)

summary(model)

```

```{r}
mcfadden_r2 <- 1 - (model$deviance / model$null.deviance)
round(mcfadden_r2, 3)
```

McFadden's pseudo R² of 0.077 indicates the model explains approximately 7.7% of the variance in cancellation outcomes. While this appears low, values between 0.02 and 0.40 are considered acceptable in logistic regression, particularly when predicting complex human behavior such as booking cancellations.

### 5.4 Model Interpretation

We convert the log odds coefficients to odds ratios using `exp()` to make them easier to interpret. An odds ratio above 1 means the variable increases cancellation risk, while an odds ratio below 1 means it decreases cancellation risk.
```{r}
round(exp(coef(model)), 3)
```

The odds ratios tell us the following:

- **lead_time**: Odds ratio of 1.005, meaning for every additional day of lead time the odds of cancellation increase by 0.5%. This compounds significantly over hundreds of days, confirming our finding in Section 4.3.
- **deposit_typeNon Refund**: Odds ratio of 40.07 — by far the strongest predictor. Bookings with non-refundable deposits are 40 times more likely to be cancelled than no-deposit bookings. This is counterintuitive but likely reflects speculative bookings made far in advance through online travel agents.
- **deposit_typeRefundable**: Odds ratio of 1.658, a moderate increase in cancellation risk compared to no-deposit bookings.
- **hotelresort hotel**: Odds ratio of 0.669, meaning resort hotel bookings are 33% less likely to be cancelled than city hotel bookings.
- **total_guests**: Odds ratio of 1.236, meaning larger bookings have slightly higher cancellation risk.
- **stays_in_week_nights**: Odds ratio of 1.065, meaning longer stays are slightly more likely to be cancelled.
- **customer_typeTransient**: Odds ratio of 2.826, meaning transient customers are nearly 3 times more likely to cancel than contract customers.
- **customer_typeGroup**: Odds ratio of 0.680, meaning group bookings are actually less likely to cancel than contract customers.
- **customer_typeTransient-Party**: Not statistically significant (p = 0.43), meaning we cannot conclude this customer type differs meaningfully from contract customers.

## Section 6: Key Insights and Recommendations

In this section we summarize the most important findings from our analysis and translate them into concrete recommendations for hotel managers.

### 6.1 Key Findings

| Finding | Detail |
|---------|--------|
| Overall cancellation rate | 27.5% of all bookings were cancelled |
| Strongest cancellation predictor | Non-refundable deposit bookings are 40x more likely to cancel |
| Lead time effect | Longer lead times significantly increase cancellation risk |
| Hotel type difference | Resort hotels have 33% lower cancellation odds than city hotels |
| Customer type risk | Transient customers are nearly 3x more likely to cancel than contract customers |
| Seasonal pricing | ADR peaks in July and August, drops in January and February |
| Lead time difference | Resort hotels have slightly longer mean lead times than city hotels |
| Family stays | Family bookings have significantly longer stays than non-family bookings |

### 6.2 Recommendations for Hotel Managers

**1. Rethink the Non-Refundable Deposit Strategy**

Non-refundable bookings have the highest cancellation rate in the dataset. While the hotel retains the payment on these cancellations, the room remains empty and ancillary revenue from food, beverage, and services is lost. Additionally, high cancellation rates on online travel agent platforms can negatively impact a hotel's search ranking and visibility. Hotels should investigate whether non-refundable policies are attracting speculative bookings and consider adjusting their distribution channel strategy accordingly.

**2. Flag High Risk Bookings Early**

Bookings with long lead times and transient customer types have the highest cancellation risk according to our model. Hotels should implement an early warning system to identify these bookings and follow up with guests closer to the arrival date.

**3. Targeted Pricing for Peak Season**

ADR peaks in July and August. Hotels should implement dynamic pricing strategies that capitalize on summer demand while offering incentives for shoulder season bookings in January and February to smooth out demand.

**4. Family Packages for Resort Hotels**

Family bookings stay significantly longer than non-family bookings. Resort hotels in particular should develop targeted family packages such as extended stay discounts and family room bundles to attract and retain this valuable segment.

**5. Focus Retention Efforts on City Hotels**

City hotels have higher cancellation rates and higher ADR than resort hotels. Reducing cancellations in city hotels would have a disproportionate impact on revenue. Loyalty programs and flexible rebooking policies could help retain these guests.

---

### 6.3 Limitations

- Duplicate rows were removed based on the assumption that identical records across all 32 columns represent data entry errors. This assumption cannot be fully verified.
- The dataset covers July 2015 to August 2017 only and may not reflect current booking behavior post-pandemic.
- The logistic regression model explains approximately 7.7% of cancellation variance, suggesting there are important predictors not captured in this dataset such as pricing changes, external events, and guest reviews.
